---
title: "How do prescription patterns for SSRIs and SNRIs vary between exam and non-exam periods in university health board regions from 2019 to 2023, and what trends are observed over this period? "
author: "Syakirah"
date: "2024-11-01"
output: html_document
---

```{r}
library(tidyverse)
library(janitor) # cleaning data
library(gt) # tables
library(here) # directory structure (will be useful later)
library(patchwork)
library(sf)
# library(rlang)
```

Read the population data and Health Board names datasets, clean names and select the relevant columns
```{r}
 population_data <- read_csv("C:/Data_Science/B223640/data/UV103_age_health_board_census.csv", skip = 10) %>%
  rename(hb_name = "Health Board Area 2019",
         hb_population = Count) %>% 
  filter(Age == "All people" & Sex == "All people") %>% 
  select(hb_name, hb_population) %>% 
  mutate(hb_name = paste("NHS", hb_name)) %>% 
  arrange(desc(hb_population))


HB_names <- read_csv("C:/Data_Science/B223640/data/HB_names.csv") %>%
  clean_names() %>% 
  select(hb, hb_name) %>% 
  rename(hbt = hb)
```

loading the age data and finding the population number of ages 18-25(typical age of university students) for each health board. Join with the population data frame to find the percentage of people in the age group within each Healthboard.

```{r}

library(dplyr)

age_data <- read.csv("C:/Data_Science/B223640/data/UV103_age_health_board_census.csv", skip = 10)

age_data <- age_data %>% 
  filter(str_detect(Age, "18|19|20|21|22|23|24|25")) %>% 
  rename(hb_name = Health.Board.Area.2019) %>% 
  mutate(hb_name = paste("NHS", hb_name))%>% 
  group_by(hb_name) %>% 
  summarise(totalpeople = sum(Count)) %>% 
  left_join(population_data) %>% 
  mutate("percentage" = totalpeople*100/hb_population)
```

Loading the exam and nonexam season datasets 

```{r}
library(tidyverse)

# List all CSV files in the "summer_20-23" folder
examfiles <- list.files("C:/Data_Science/B223640/data/examszn_19-23", pattern = "csv", full.names = TRUE)

column_types <- cols(
  HBT = col_character(),
  DMDCode = col_character(), # Ensures DMDCode is read as character in all files
  BNFItemCode = col_character(),
  BNFItemDescription = col_character(),
  PrescribedType = col_character(),
  GPPractice = col_double(),
  NumberOfPaidItems = col_double(),
  PaidQuantity = col_double(),
  GrossIngredientCost = col_double(),
  PaidDateMonth = col_double()
)

examszndata <- examfiles %>% 
  map_dfr(~read_csv(., col_types = column_types))

nonexamfiles <- list.files("C:/Data_Science/B223640/data/nonexamszn_19-23", pattern = "csv", full.names = TRUE)

nonexamszndata <- nonexamfiles %>% 
  map_dfr(~read_csv(., col_types = column_types))
```

joining the the exam and exam season datasets with the HB_names and population_data sets and filter for the SSRIs and SNRIs commonly prescribed for exam-stress related anxiety and depression. the exam season data was filtered to include only rows for University health board regions(the 6 healthboard with the highest percentage of persons 18-25 years of age). This would help narrow the scope to only investigate the trends where young persons are more prevalent.


```{r}
examszndata1 <- examszndata %>%
  clean_names() %>% 
  full_join(HB_names) %>%
  full_join(population_data) %>%  
  filter(str_detect(bnf_item_description, "ESCITALOPRAM|SERTRALINE|FLUOXETINE|VENLAFAXINE|FLUVOXAMINE|PAROXETINE|CITALOPRAM|RITALIN|CONCERTA"))%>%
  filter(str_detect(hb_name, "Lothian|Grampian|Greater Glasgow and Clyde|Tayside|Fife|Forth Valley")) %>% 
  group_by(hb_name, bnf_item_description, hb_population) %>%
  summarise(total_prescription = sum(paid_quantity)) %>% 
  mutate("per_person" =  total_prescription/hb_population)

nonexamszndata1 <- nonexamszndata %>%
  clean_names() %>% 
  full_join(HB_names) %>% 
  full_join(population_data) %>%  
  filter(str_detect(bnf_item_description, "ESCITALOPRAM|SERTRALINE|FLUOXETINE|VENLAFAXINE|FLUVOXAMINE|PAROXETINE|CITALOPRAM|RITALIN|CONCERTA"))%>%
   filter(str_detect(hb_name, "Lothian|Grampian|Greater Glasgow and Clyde|Tayside|Fife|Forth Valley")) %>%
  group_by(hb_name, bnf_item_description, hb_population) %>%
  summarise(total_prescription = sum(paid_quantity)) %>% 
  mutate("per_person" =  total_prescription/hb_population)

```
Pivot wider to find the individual total prescriptions of each medication for each Health Board
```{r}
library(dplyr)
library(stringr)

 examszndatawide <- examszndata %>%
   clean_names() %>% 
  mutate(bnf_item_description = str_remove(bnf_item_description, "[_\\s].*")) %>% 
   full_join(HB_names) %>% 
  full_join(population_data) %>%  
  filter(str_detect(bnf_item_description, "ESCITALOPRAM|SERTRALINE|FLUOXETINE|VENLAFAXINE|FLUVOXAMINE|PAROXETINE|CITALOPRAM|RITALIN|CONCERTA"))%>%
   filter(str_detect(hb_name, "Lothian|Grampian|Greater Glasgow and Clyde|Tayside|Fife|Forth Valley")) %>%
  group_by(hb_name, bnf_item_description, hb_population) %>%
  summarise(total_prescription = sum(paid_quantity)) %>% 
   pivot_wider(
     names_from = bnf_item_description,
     values_from = total_prescription,
     values_fill = list(total_prescription = 0)
   ) %>% 
   left_join(age_data) %>% 
   mutate(average_prescription = rowMeans(across(CITALOPRAM:VENLAFAXINE), na.rm = TRUE)) %>% 
   arrange(desc(average_prescription)) %>% 
   gt()
 
nonexamszndatawide <- nonexamszndata %>%
   clean_names() %>% 
  mutate(bnf_item_description = str_remove(bnf_item_description, "[_\\s].*")) %>% 
   full_join(HB_names) %>% 
  full_join(population_data) %>%  
  filter(str_detect(bnf_item_description, "ESCITALOPRAM|SERTRALINE|FLUOXETINE|VENLAFAXINE|FLUVOXAMINE|PAROXETINE|CITALOPRAM|RITALIN|CONCERTA"))%>%
   filter(str_detect(hb_name, "Lothian|Grampian|Greater Glasgow and Clyde|Tayside|Fife|Forth Valley")) %>%
  group_by(hb_name, bnf_item_description, hb_population) %>%
  summarise(total_prescription = sum(paid_quantity)) %>% 
   pivot_wider(
     names_from = bnf_item_description,
     values_from = total_prescription,
     values_fill = list(total_prescription = 0)
   ) %>% left_join(age_data) %>% 
  mutate(average_prescription = rowMeans(across(CITALOPRAM:VENLAFAXINE), na.rm = TRUE)) %>% 
   arrange(desc(average_prescription))
```


```{r}

examseasontry <- examszndata1 %>%
    mutate(bnf_item_description = str_remove(bnf_item_description, "[_\\s].*")) %>%
  full_join(age_data) %>% 
  group_by(hb_name, totalpeople, percentage, hb_population) %>% 
  summarise(total_prescription = sum(total_prescription)) %>% 
  filter(!is.na(total_prescription)) %>% 
  mutate(Season = "examseason") %>% 
  mutate(per_person = total_prescription/hb_population)

nonexamseasontry <- nonexamszndata1 %>%
    mutate(bnf_item_description = str_remove(bnf_item_description, "[_\\s].*")) %>%
  full_join(age_data) %>% 
  group_by(hb_name, totalpeople, percentage, hb_population) %>% 
  summarise(total_prescription = sum(total_prescription)) %>% 
  filter(!is.na(total_prescription)) %>% 
  mutate(Season = "nonexamseason")%>% 
  mutate(per_person = total_prescription/hb_population)

combined_data <- bind_rows(examseasontry, nonexamseasontry)

ggplot(combined_data, aes(x = Season, y = total_prescription)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5, color = "blue")
  labs(title = "Comparison of Total Prescriptions between Datasets",
       x = "Season",
       y = "Total Prescriptions") +
  theme_minimal()
  
```
Throughout 2019 to 2023 cumulatively, total prescriptions of SNRIs and SSRIs increase in the months when scottish universities have exams, but only slightly.

Trends of each medication could be plotted on a time series graph for each year to see if there are any patterns which indicate increase in prescription rates during exam seasons. 

Graph of total prescriptions against the total number of persons aged 18-25 in each university healthboard regions. 
```{r}

combined_data %>%
  ggplot(aes(x = totalpeople, y = total_prescription, color = Season)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +  # Adds line of best fit
  labs(title = "Total People vs. Total Prescription",
       x = "Total People",
       y = "Total Prescription",
       color = "Season") +
  theme_minimal()
```



